\section{Introduction}
\label{sec:introduction}

Machine learning has recently enabled solutions to problems that were difficult or infeasible with traditional algorithms. This progress produced both large, general-purpose models (e.g., large language models) and many compact, task-specific models.

Most deployments still rely on centralized data centres, which are efficient but impose limitations: they require network connectivity, add latency, and raise privacy concerns.

Edge AI reduces these limitations by running inference locally on user devices. TinyML \cite{wardenTinyMLMachineLearning2019} focuses on the extreme end of this spectrum, designing models that run on microcontrollers (MCUs). MCUs are severely resource-constrained: typical devices provide one or two CPU cores (usually clocked below 240 MHz) and on the order of 512 KB of RAM, so they cannot host large models or their full-precision weights. Low latency is also essential because many TinyML applications (for example, audio or video streams) require real-time processing.

Numerous model families have been developed for different trade-offs, including multilayer perceptrons (MLPs), convolutional neural networks (CNNs), recurrent neural networks (RNNs), and Transformers. The recently proposed Mamba architecture \cite{guMambaLinearTimeSequence2024} has emerged as a promising alternative to Transformers for long-context tasks. Mamba combines ideas from CNNs and RNNs and leverages Selective State-Space Model (S6) block to achieve a smaller memory footprint while maintaining strong performance on long sequences.

Prior work has explored optimizations for Mamba such as quantization \cite{xuMambaQuantQuantizingMamba2025}, pruning, and knowledge distillation \cite{shihabEfficientUnstructuredPruning}. MambaLite-Micro \cite{xuMambaLiteMicroMemoryOptimizedMamba2025} implemented PyTorch-style inference for Mamba directly in C and ran it on ESP32-S3 and STM32 microcontrollers; this preserves compatibility but makes architecture changes and further optimizations cumbersome.

This paper addresses the following overarching research question:
\textit{How can Mamba architectures be adapted and optimized for efficient inference on resource-constrained microcontrollers?}
To address this comprehensively, we formulate three specific sub-questions:
\begin{itemize}
  \item \textbf{RQ1:} What are the primary memory and computational bottlenecks when deploying Mamba models on microcontrollers?
  \item \textbf{RQ2:} How do standard quantization strategies affect Mamba's inference latency, memory footprint, and task accuracy on these edge devices?
  \item \textbf{RQ3:} To what extent can hardware-unfriendly components of the Mamba architecture (e.g., specific activation functions) be simplified or replaced to fit microcontroller constraints, and what are the resulting trade-offs in performance?
\end{itemize}
To answer these questions, we profile deployment strategies on actual hardware, isolate the underlying hardware bottlenecks, and evaluate the efficacy of translating typically GPU-centric optimizations to microcontroller architectures.

The remainder of this paper is organized as follows. Section~\ref{sec:background} surveys related work and foundational background. Section~\ref{sec:methodology} details our proposed optimizations and architectural modifications, while Section~\ref{sec:experimental_setup} outlines how the experiments were conducted. Experimental results and a subsequent discussion are presented in Section~\ref{sec:results_and_discussion}. Finally, Section~\ref{sec:responsible_research} addresses ethical considerations, and Section~\ref{sec:conclusion} concludes the paper with directions for future work.
