\section{Responsible Research}
\label{sec:responsible_research}

This section outlines the ethical considerations, reproducibility frameworks, and the structured use of generative AI tools throughout this study. Specifically, we document dataset and code compliance, experimental replication protocols, and potential dual-use risks alongside corresponding mitigation strategies.

\subsection{Ethical Considerations}

While this study does not involve human subjects or direct third-party participants, the broader implications of deploying machine learning on edge devices warrant careful ethical consideration. By optimizing machine learning frameworks for resource-constrained microcontrollers, this work shifts computational workloads from centralized datacenters to low-power edge devices, thereby contributing to environmental sustainability via reduced operational carbon footprints.

Furthermore, on-device processing inherently enhances user privacy by keeping raw data local and minimizing cloud-based data exposure. However, decentralized processing also introduces potential risks, such as facilitating covert local surveillance. Consequently, downstream deployments must integrate robust hardware safeguards, operational transparency, and appropriate governance frameworks. Because TinyML applications frequently process highly sensitive telemetry (e.g., biomedical or acoustic data), developers utilizing this framework must implement stringent security measures and maintain transparency regarding data lifecycle management.

Additionally, the proliferation of specialized edge devices introduces e-waste risks due to rapid hardware obsolescence. To mitigate this impact, practitioners are encouraged to prioritize hardware longevity, provide long-term software maintenance, and adhere to responsible electronic waste disposal practices.

Finally, we acknowledge the systematic risk of algorithmic and dataset bias. The datasets benchmarked in this work (HAR and Google Speech Commands) are established standards within the TinyML community; however, they may exhibit representation gaps across diverse populations or operational environments. These limitations must be factored into the generalization and interpretation of the empirical results.

\subsection{Reproducibility}

All datasets utilized in this research are publicly available, and all associated software dependencies are governed by open-source licenses permitting research usage. To ensure reproducibility, we have comprehensively documented the experimental pipeline, including model training, optimization, quantization, and microcontroller benchmarking. While full replication requires specific hardware targets, the ESP32 family utilized in this study is widely accessible within the research community.

To ensure empirical fidelity, performance profiling was conducted using a unified, consistent profiler across all experimental iterations, with results systematically annotated by the specific microcontroller variant used. The complete codebase—including replication scripts, library dependencies, and exact version constraints—is hosted in a public repository to facilitate community verification.

\subsection{Use of AI Tools}

Generative AI tools were utilized strictly as supportive instruments during the development and drafting phases of this research. AI assistance was restricted to writing utility scripts, debugging software implementations, and refining the grammatical clarity of the manuscript.

Specifically, AI tools were leveraged for the following tasks:
\begin{itemize}
  \item Automating boilerplate scripts for the experimental pipeline execution.
    \author
  \item Debugging syntax and integration issues within the experimental codebase.
  \item Enhancing the syntactic clarity, flow, and structure of the written narrative.
  \item Assisting with automated \LaTeX{} formatting and structural consistency.
\end{itemize}

All core scientific decisions, experimental designs, and data interpretations were executed independently by the authors. Generative AI tools were not employed to synthesize, alter, or influence experimental outcomes, empirical data, or scientific conclusions.