\section{Conclusions and Future Work}
\label{sec:conclusion}

This paper examined how Mamba can be adapted and optimized for inference on microcontrollers. Our experiments support three concise findings:

\begin{itemize}
  \item Deploying via TFLM (with model-graph stripping) substantially reduces peak RAM compared to the C-based MambaLite-Micro workflow (roughly a 75\% reduction in our runs).
  \item Memory vs. latency trade-off: INT8 quantization cuts peak RAM by about $\approx33\%$ and model file size by $\approx60\%$, but on the tested ESP32-class hardware it increased latency because efficient INT8 vector instructions are lacking.
  \item Split vs. full deployment: splitting into pre-SSM / SSM-step / post-SSM lowers peak RAM and permits longer sequences, but increases sensitivity to static quantization and complicates the runtime pipeline.
\end{itemize}

Given our findings, the TFLM-based deployment pipeline seems to be a better path for Mamba on microcontrollers than the C-based approach of MambaLite-Micro, as it allows for more flexible experimentation and even better memory efficiency. While the C-based approach could allow for even more aggressive optimizations they would be hard to deploy for general use models. To make the deployment more practical the model split that we implemented would need to be integrated into the TFLM as a custom operator.

Taken together, Mamba is a strong candidate for TinyML on MCUs when memory optimizations are prioritized, but real-world latency depends heavily on hardware and kernel support. A next step that could be pursued is implementing a custom, quantized Selective State-Space Model kernel for TFLM and running targeted microbenchmarks across MCU families to guide low-level optimizations. The model split that we applied was applied in order to mitigate the lack of a looping mechanism in TFLM, but it had a side effect of reducing the peak RAM usage, the reason fo this could be further investigated.
We also recommend systematically evaluating pruning methods (structured and unstructured) to reduce parameter count and memory footprint, and combining those experiments with Quantization-Aware Training and lower-bit formats.
